Notes on conformal manifolds

### 2025-09-25 Draft_MLFlatDirections.tex
Modify the draft, beginning of section 4 on discussion of IS/AIS/Sequential MC AND add the modifications for the figures generation

	To address these challenges, we formulate the search for annihilating polynomials as a determination of a density probability on the space of polynomials that stresses higher probabilities on polynomials $p$ that minimises $\displaystyle \sum_{i=1}^{n_{\rm points}}p(\vec{x}^{(i)})^{2}$. Such a probability density is \textit{a priori} hard to find, but can be approximated using variations of Important Sampling methods~\cite{910e8e7d-2e70-317c-adf6-e21cd50755a1}. Importance sampling is a method to parametrise expectation values with respect to a target distribution $d$, from a weighted average using another distribution $\pi$. It is based on the observation that the expectation
	\begin{equation}
		\mathbb{E}_{d}[f] = \int f(z)d(z)\ \d z
	\end{equation}
	of a function $f$ with respect to $d$ can be computed from the expectation $\mathbb{E}_{\pi}[w f]$, now computed with respect to the distribution $\pi$, with unnormalised $w(z) = d(z)/\pi(z)$. Therefore, if it is easier to draw a sample $\{z^{(k)}\}_{k\in\llbracket 1, n_{\rm particles}\rrbracket}$ from $\pi$ than from $d$, one can estimate $\mathbb{E}_{d}[f]$ from the weighted average
	\begin{equation}
		\mathbb{E}_{d}[f] \simeq \frac{\displaystyle \sum_{k=1}^{n_{\rm particles}} w(z^{(k)})\,f(z^{(k)})}{\displaystyle \sum_{i=1}^{n_{\rm particles}} w(z^{(k)})}.
	\end{equation}
	The accuracy of this estimation depends on the size of the sample and, crucially, on the choice of the distribution $\pi$. \ce{Add conclusion saying that you can parametrise $d$ using a sample and weights, and eq. as (4.4).}

	To have an accurate estimation of $d$, we employ an Annealed Importance Sampling (AIS) \cite{neal1998annealedimportancesampling} combined with Sequential Monte Carlo (SMC) methods \cite{del2006sequential,naesseth2024elementssequentialmontecarlo}. AIS aims at estimating $d$ sequentially from multiple intermediate distributions $\pi_{n}$ that smoothly transition from an initial, tractable distribution $d_{0}$ (chosen to be easy to sample from) to the complex target distribution $d$.
	These distributions are approximated by weighted samples $\{z^{(k)}_{n},w^{(k)}_{n}\}$ of particles, each representing a candidate symbolic expression, such that the resulting weighted average estimates $\pi_{n}$: \ce{Normalise the weights?}
	\begin{equation}
		\frac{\displaystyle \sum_{k=1}^{n_{\rm particles}} w^{(k)}_{n}\,f(z^{(k)}_{n})}{\displaystyle \sum_{k=1}^{n_{\rm particles}} w^{(k)}_{n}} \underset{n_{\rm particles}\rightarrow+\infty}{\longrightarrow} \mathbb{E}_{\pi_{n}}[f],
	\end{equation}
	or equivalently
	\begin{equation} \label{eq:pifromweights}
		\pi_{n}(z) \simeq \frac{\displaystyle \sum_{k=1}^{n_{\rm particles}} w^{(k)}_{n}\,\delta(z-z^{(k)}_{n})}{\displaystyle \sum_{k=1}^{n_{\rm particles}} w^{(k)}_{n}}.
	\end{equation}
	The $\pi_{n}$ are incrementally deformed using a Markov kernel to gradually reproduce the target $d$ and the particles get reweighed. This annealing process is guided by a temperature-like parameter that gradually emphasises the data likelihood, allowing for more efficient and tunable exploration of the polynomials landscape.

	This procedure has the disadvantage that the weights variance tends to increase, leading to weight degeneracy. We use SMC methods to incorporate a resampling procedure to AIS, thus focussing computational effort on high-probability regions.  This combination of importance sampling, mutation and resampling maintains diversity among the particles and prevents premature convergence to suboptimal models. These features make AIS-SMC particularly well-suited for symbolic regression tasks, where the search space is not only high-dimensional but also structured and discontinuous.

>>>>

	To address these challenges, we formulate the search for annihilating polynomials as a determination of a density probability on the space of polynomials that stresses higher probabilities on polynomials $p$ that minimises $\displaystyle \sum_{i=1}^{n_{\rm points}}p(\vec{x}^{(i)})^{2}$. Such a probability density is \textit{a priori} hard to find, but can be approximated using variations of Important Sampling methods~\cite{910e8e7d-2e70-317c-adf6-e21cd50755a1}. Importance sampling is a method to parametrise expectation values with respect to a target distribution $d$, for which you can compute $d(z)$ for some sample $z$, but for which you do not know how to draw from. IS uses another distribution $\pi$ whose density probabilty is proportional to $d$, but from which you know how to draw from. IS then parametrizes the density $d$ with a sequence of weights $w^{(k)}$ and is based on the observation that the expectation
	\begin{equation}
		\mathbb{E}_{d}[f] = \int f(z)d(z)\ \d z
	\end{equation}
	for some function $f$ with respect to $d$ can be computed from the expectation $\mathbb{E}_{\pi}[w f]$, now computed with respect to the distribution $\pi$, with unnormalised $w^{(k)} = d(z^{k})/\pi(z^{(k)})$. Therefore, if it is easier to draw a sample $\{z^{(k)}\}_{k\in\llbracket 1, n_{\rm particles}\rrbracket}$ from $\pi$ than from $d$, one can estimate $\mathbb{E}_{d}[f]$ from the weighted average
	\begin{equation}
		\mathbb{E}_{d}[f] \simeq \displaystyle \sum_{k=1}^{n_{\rm particles}} w(z^{(k)})\,f(z^{(k)})/\displaystyle \sum_{i=1}^{n_{\rm particles}} w(z^{(k)}).
	\end{equation}
	or stated differently 
	\begin{equation} \label{eq:}
		d(z) \simeq \displaystyle \sum_{k=1}^{n_{\rm particles}} w^{(k)}\,\delta(z-z^{(k)})/\displaystyle \sum_{k=1}^{n_{\rm particles}} w^{(k)}.
	\end{equation}	
	The accuracy of this estimation depends on the size of the sample and, crucially, on the choice of the distribution $\pi$.

	However, providing a good $\pi$ distribution may still remains an hard problem. In order to fix this point, we employ a so called Annealed Importance Sampling (AIS) \cite{neal1998annealedimportancesampling}. AIS aims at estimating $\pi$ sequentially from multiple intermediate distributions $\pi_{n}$ that smoothly transition from an initial, tractable distribution $\pi_{0}$ (chosen to be easy to sample from) to the complex target distribution $\pi$.
	These distributions are approximated by weighted samples $\{z^{(k)}_{n},w^{(k)}_{n}\}$ of particles, such that the resulting weighted average estimates $\pi_{n}$
	\begin{equation}
		\mathbb{E}_{\pi_{n}}[f] \simeq \displaystyle \sum_{k=1}^{n_{\rm particles}} \tilde{w}^{(k)}_{n}\,f(z^{(k)}_{n})
	\end{equation}
	or equivalently
	\begin{equation} \label{eq:pifromweights}
		\pi_{n}(z) \simeq \displaystyle \sum_{k=1}^{n_{\rm particles}} \tilde{w}^{(k)}_{n}\,\delta(z-z^{(k)}_{n}).
	\end{equation}
	with  $\tilde{w}^{(k)}_n  = w^{(k)}_n/\sum_{k=1}^{n_{\rm particles}} w^{(k)}_{n}$ the normalised weights. The $\pi_{n}$ are incrementally deformed using a Markov kernel to gradually reproduce the target $d$ and the particles get reweighed. This annealing process is guided by a temperature-like parameter that gradually emphasises the data likelihood, allowing for more efficient and tunable exploration of the polynomials landscape. The approximation procedure can be sum up as follow 
	\begin{equation}
	\pi_0(z) \underset{\rm Annealing}{\longrightarrow}\pi_n(z) \simeq \displaystyle \sum_{k=1}^{n_{\rm particles}} \tilde{w}^{(k)}_{n}\,\delta(z-z^{(k)}_{n}) \underset{\rm{n} \rightarrow +\infty}{\longrightarrow} \pi(z) \simeq \displaystyle \sum_{k=1}^{n_{\rm particles}} \tilde{w}^{(k)}\,\delta(z-z^{(k)}) \underset{\rm{n_{particles}} \rightarrow +\infty}{\longrightarrow} d(z)
	\nonumber
	\end{equation}
	Therefore, we start from an intial, easy to sample distribution $\pi_0$, that is parametrised by a set of particles and weights $\{z_0^{(k)},w_0^{(k)}\}$. We gradually mutate them with an annealing procedure to get a set of $\{z_n^{(k)},w_n^{(k)}\}$, which are used to parametrise $\pi_n$. As $n$ increases, the procedure converges to the distribution $\pi$, which itself is used to parametrise the target distribution $d$.

	This procedure has the disadvantage that the weights variance tends to increase, leading to weight degeneracy. We combine the previously mentionne AIS technique with SMC methods  \cite{del2006sequential,naesseth2024elementssequentialmontecarlo} to incorporate a resampling procedure, thus focussing computational effort on high-probability regions.  This combination of importance sampling, mutation and resampling maintains diversity among the particles and prevents premature convergence to suboptimal models. These features make AIS-SMC particularly well-suited for symbolic regression tasks, where the search space is not only high-dimensional but also structured and discontinuous.
